---
title: "Bayesian analysis"
bibliography: "../../Literature/library.bib"
---

## Data analysis overview

```{r prep, include=FALSE}
# When working interactively (i.e. not calling this file from another
# function e.g. 'Knit' button or render() in outside script), set working
# directory to the location of this file. This way you will be working from
# the same perspective as those outside functions and paths will work
# either way.
# getwd()
# setwd("./source/R-new")

# Load packages and our functions
# Plot formatting
source("AMgraph.R")
# Our curve functions
source("AMfunctions.R")
# Interface to WinBUGS
require(R2WinBUGS)
require(lattice)
require(car)
data.path <- "../data" 

options(width=90, scipen = 4, show.signif.stars = FALSE)
# Some of the graphical model checking steps use different plotting
# parameters, so we keep the original plotting parameters to re-load when
# we're done
orig.par <- par()
```

 
In this tutorial, we will use Markov Chain Monte Carlo (MCMC) methods and Bayesian inference to estimate the variance model and the curve function parameters in one step. R will send the data and instructions to Bayesian software, WinBUGS, and then retreive the results for display using the R package, `R2WinBUGS`. If you have not already installed WinBUGS, the software download and instructions are found [here](http://www.mrc-bsu.cam.ac.uk/software/bugs/the-bugs-project-winbugs/). (Very similar implementations are possible with OpenBUGS or JAGS and their respective R packages.)

## Preparatory steps

First we reload the O'Connell ELISA data as in previous tutorials:

```{r ocon-entry}
# O'Connell ELISA data 
ocon <- read.csv(file.path(data.path, "ocon.csv"), header = TRUE)
head(ocon)
```

```{r model1}
# Simple 4pl model: Constant variance (as standard deviation, sigma) ----
# Need number of observations
J <- nrow(ocon)
# all the data minus the rep variable
data <- c(J = J, as.list(ocon[ , c(1, 3)]))
#str(data)

# Initial values
# Could be initial values from the classical model
#inits0 <- as.numeric(y.coef)
# Or simply choose reasonable values
inits1 <- c(1, 0.1, 3000, -1, 0.002)
names(inits1) <- c("beta1", "beta2", "beta3", "beta4", "var")
# If running more than 1 chain, need a set of initial values for each chain
# Maybe as a function such as adding randomness
# inits2 <- 
# inits3 <- 
# As with other BUGS inputs, format as a list
inits <- list(inits1
              #, inits2, inits3
              )

# Parameters to monitor (only need to monitor betas and sigma this time)
parameters <- c(#"x0", 
                "beta1", "beta2", "beta3", "beta4", "sigma"
                )

model1 <- function(){
    for (i in 1:J){
        mu[i] <- beta2 + (beta1 - beta2)/(1 + pow(conc[i] / beta3, beta4))
        od[i] ~ dnorm(mu[i], tau)
        # Original code:
        # Is this right? sigma[i] but one tau?
        # sigma[i] <- 1.0/(sqrt(tau))
    }
    # Should not allow betas 1-3 to be <= 0 
    # (we know that even at x=0, y ne 0 unless it was transformed)
    # beta1 and beta2 had dnorm(0, 0.001), but a mean of zero is not
    # realistic. Are dgammas OK for beta1 and beta2? Or log-normal?
    # Also, beta1 must be > beta2. Could this constraint be added?
    beta1 ~ dgamma(0.001,0.001)
    beta2 ~ dgamma(0.001,0.001)
    beta3 ~ dgamma(0.001,0.001)
    # We know that with an immunometric assay beta4 must be <= 0.     
    # Should it be specified like this?
    beta4 ~ dunif(-10,0)
    # Precision, tau, cannot be negative
    # tau ~ dgamma(0.0001,0.001)
    # Would like to monitor sigma rather than tau:
    tau <- 1/var
    var ~ dgamma(0.001,0.001)
    sigma <- sqrt(var)
}
write.model(model1)
bugs.out <- bugs(data, inits = inits, parameters, "model.bug",
                 n.chains = 1, n.iter = 2000, 
                 bugs.directory="c:/Program Files/WinBUGS14/"
                 , working.directory=getwd()
                 , debug = F
)
# Parameter posterior distributions
mcmc.out1 <- as.mcmc.list(bugs.out)
summary(mcmc.out1, q = c(0.025, 0.975))

# Nice plots
xyplot(mcmc.out1)
densityplot(mcmc.out1)
#autocorr.plot(mcmc.out1) # not super clear (remember how to interpret)

```


```{r model2}
# Simple 4pl model: Constant variance (as standard deviation, sigma) ----
# Everything is the same as model 1 except the priors on betas 1--3
model2 <- function(){
    for (i in 1:J){
        mu[i] <- beta2 + (beta1 - beta2)/(1 + pow(conc[i] / beta3, beta4))
        od[i] ~ dnorm(mu[i], tau)
        # Original code:
        # Is this right? sigma[i] but one tau?
        # sigma[i] <- 1.0/(sqrt(tau))
    }
    # Should not allow betas 1-3 to be <= 0 
    # (we know that even at x=0, y ne 0 unless it was transformed)
    # beta1 and beta2 had dnorm(0, 0.001), but a mean of zero is not
    # realistic. Are dgammas OK for beta1 and beta2? Or log-normal?
    # Also, beta1 must be > beta2. Could this constraint be added?
    beta1 ~ dunif(0, 10)
    beta2 ~ dunif(0, 10)
    beta3 ~ dunif(0, 100000) # we really don't know a priori...
    # We know that with an immunometric assay beta4 must be <= 0.     
    # Should it be specified like this?
    beta4 ~ dunif(-10,0)
    # Precision, tau, cannot be negative
    # tau ~ dgamma(0.0001,0.001)
    # Would like to monitor sigma rather than tau:
    tau <- 1/var
    var ~ dgamma(0.001,0.001)
    sigma <- sqrt(var)
}
write.model(model2)
bugs.out <- bugs(data, inits = inits, parameters, "model.bug",
                 n.chains = 1, n.iter = 2000, 
                 bugs.directory="c:/Program Files/WinBUGS14/"
                 , working.directory=getwd()
                 , debug = F
)
# Parameter posterior distributions
mcmc.out2 <- as.mcmc.list(bugs.out)
summary(mcmc.out2, q = c(0.025, 0.975))

# Nice plots
xyplot(mcmc.out2)
densityplot(mcmc.out2)
#autocorr.plot(mcmc.out2) # not super clear (remember how to interpret)

```

```{r model3}
# Heteroscedastic 4pl model: variance changes as a function of y ----
# Data is the same as model 1 and 2
# Initial values for variance model change
# Chose reasonable values
inits1 <- c(1, 0.1, 3000, -1, 0.002, 1.2)
names(inits1) <- c("beta1", "beta2", "beta3", "beta4", "sigma.sq", "theta")
# If running more than 1 chain, need a set of initial values for each chain
# Maybe as a function such as adding randomness
# inits2 <- 
# inits3 <- 
# As with other BUGS inputs, format as a list
inits <- list(inits1
              #, inits2, inits3
              )

# Parameters to monitor (only need to monitor betas and sigma this time)
parameters <- c(#"x0", 
                "beta1", "beta2", "beta3", "beta4", "sigma", "theta"
                )
model3 <- function(){
    for (i in 1:J){
        mu[i] <- beta2 + (beta1 - beta2)/(1 + pow(conc[i] / beta3, beta4))
        # Tau changes
        od[i] ~ dnorm(mu[i], tau[i])
        tau[i] <- 1/var[i]
        # same variance model from previous tutorials 
        var[i]  <- sigma.sq * pow(mu[i], 2*theta)
    }
    # Should not allow betas 1-3 to be <= 0 
    # (we know that even at x=0, y ne 0 unless it was transformed)
    # beta1 and beta2 had dnorm(0, 0.001), but a mean of zero is not
    # realistic. Are dgammas OK for beta1 and beta2? Or log-normal?
    # Also, beta1 must be > beta2. Could this constraint be added?
    beta1 ~ dunif(0, 10)
    beta2 ~ dunif(0, 10)
     # reduced the upper limit x10 because would not converge
    beta3 ~ dunif(0, 10000)
    # We know that with an immunometric assay beta4 must be <= 0.     
    # Should it be specified like this?
    beta4 ~ dunif(-10,0)
    # Precision, tau, cannot be negative
    # Would like to monitor sigma rather than tau or var, but advice on priors
    # for variance are more common. And for consistency with previous constant
    # variance models (model2):
    sigma.sq ~ dgamma(0.001,0.001)
    sigma <- sqrt(sigma.sq)
    # Could use this prior for theta
    theta ~ dunif(0, 5)
    # Or could give stronger prior belief to constant variance (ie. theta = 0)
    # and not exclude possibility of decreasing variance...
    # theta ~ dnorm(0, 0.1)
}
write.model(model3)
bugs.out <- bugs(data, inits = inits, parameters, "model.bug",
                 n.chains = 1, n.iter = 2000, 
                 bugs.directory="c:/Program Files/WinBUGS14/"
                 , working.directory=getwd()
                 , debug = T
)
# Parameter posterior distributions
mcmc.out3 <- as.mcmc.list(bugs.out)
summary(mcmc.out3, q = c(0.025, 0.975))

# Nice plots
xyplot(mcmc.out3)
densityplot(mcmc.out3)
#autocorr.plot(mcmc.out2) # not super clear (remember how to interpret)

```

```{r model4}
# Heteroscedastic 4pl model: variance changes as a function of y ----
# Everything is the same as model3 except prior for theta

model4 <- function(){
    for (i in 1:J){
        mu[i] <- beta2 + (beta1 - beta2)/(1 + pow(conc[i] / beta3, beta4))
        # Tau changes
        od[i] ~ dnorm(mu[i], tau[i])
        tau[i] <- 1/var[i]
        # same variance model from previous tutorials 
        var[i]  <- sigma.sq * pow(mu[i], 2*theta)
    }
    # Should not allow betas 1-3 to be <= 0 
    # (we know that even at x=0, y ne 0 unless it was transformed)
    # beta1 and beta2 had dnorm(0, 0.001), but a mean of zero is not
    # realistic. Are dgammas OK for beta1 and beta2? Or log-normal?
    # Also, beta1 must be > beta2. Could this constraint be added?
    beta1 ~ dunif(0, 10)
    beta2 ~ dunif(0, 10)
     # reduced the upper limit x10 because would not converge
    beta3 ~ dunif(0, 10000)
    # We know that with an immunometric assay beta4 must be <= 0.     
    # Should it be specified like this?
    beta4 ~ dunif(-10,0)
    # Precision, tau, cannot be negative
    # Would like to monitor sigma rather than tau or var, but advice on priors
    # for variance are more common. And for consistency with previous constant
    # variance models (model2):
    sigma.sq ~ dgamma(0.001,0.001)
    sigma <- sqrt(sigma.sq)
    # Could use this prior for theta
    # theta ~ dunif(0, 5)
    # Or could give stronger prior belief to constant variance (ie. theta = 0)
    # and not exclude possibility of decreasing variance...
    theta ~ dnorm(0, 0.1) 
}
write.model(model4)
bugs.out <- bugs(data, inits = inits, parameters, "model.bug",
                 n.chains = 1, n.iter = 2000, 
                 bugs.directory="c:/Program Files/WinBUGS14/"
                 , working.directory=getwd()
                 , debug = F
)
# Parameter posterior distributions
mcmc.out4 <- as.mcmc.list(bugs.out)
summary(mcmc.out4, q = c(0.025, 0.975))

# Nice plots
xyplot(mcmc.out4)
densityplot(mcmc.out4)
#autocorr.plot(mcmc.out2) # not super clear (remember how to interpret)

```

Model is slower to converge with this prior for theta, but estimates are similar.

```{r model4-newyimp}
# Heteroscedastic 4pl model: variance changes as a function of y ----
# plus one new od observation (unknown conc)
new.od <- c(conc = NA, rep = 4, od = 0.5)
ocon2 <- rbind(ocon, new.od)
# Recount number of observations
J2 <- nrow(ocon)
# all the data minus the rep variable
data <- c(J = J2, as.list(ocon[ , c(1, 3)]))
#str(data)
# Initial values for variance model change
# Chose reasonable values
inits1 <- c(1, 0.1, 3000, -1, 0.002, 1.2
            #, 3, 1
            )
names(inits1) <- c("beta1", "beta2", "beta3", "beta4", "sigma.sq", "theta"
                   #, "mu.x", "sd.x"
                   )
# If running more than 1 chain, need a set of initial values for each chain
# Maybe as a function such as adding randomness
# inits2 <- 
# inits3 <- 
# As with other BUGS inputs, format as a list
inits <- list(inits1
              #, inits2, inits3
              )

# Parameters to monitor (only need to monitor betas and sigma this time)
parameters <- c(#"x0", 
                "beta1", "beta2", "beta3", "beta4", "sigma", "theta"
                #, "conc[37]"
                )
model4imp <- function(){
    for (i in 1:J){
        mu[i] <- beta2 + (beta1 - beta2)/(1 + pow(conc[i] / beta3, beta4))
        # Tau changes
        od[i] ~ dnorm(mu[i], tau[i])
        tau[i] <- 1/var[i]
        # same variance model from previous tutorials 
        var[i]  <- sigma.sq * pow(mu[i], 2*theta)
        # distribution of conc[i]
        # conc[i] ~ dnorm(mu.x, p.x)
    }
    # Should not allow betas 1-3 to be <= 0 
    # (we know that even at x=0, y ne 0 unless it was transformed)
    # beta1 and beta2 had dnorm(0, 0.001), but a mean of zero is not
    # realistic. Are dgammas OK for beta1 and beta2? Or log-normal?
    # Also, beta1 must be > beta2. Could this constraint be added?
    beta1 ~ dunif(0, 10)
    beta2 ~ dunif(0, 10)
     # reduced the upper limit x10 because would not converge
    beta3 ~ dunif(0, 10000)
    # We know that with an immunometric assay beta4 must be <= 0.     
    # Should it be specified like this?
    beta4 ~ dunif(-10,0)
    # Precision, tau, cannot be negative
    # Would like to monitor sigma rather than tau or var, but advice on priors
    # for variance are more common. And for consistency with previous constant
    # variance models (model2):
    sigma.sq ~ dgamma(0.001,0.001)
    sigma <- sqrt(sigma.sq)
    # Could use this prior for theta
    # theta ~ dunif(0, 5)
    # Or could give stronger prior belief to constant variance (ie. theta = 0)
    # and not exclude possibility of decreasing variance...
    theta ~ dnorm(0, 0.1) 
    # missing value priors
#     mu.x ~ dunif(0, 10000)
#     p.x <- 1/pow(sd.x, 2)
#     sd.x ~ dunif(0, 10)
}
write.model(model4imp)
bugs.out <- bugs(data, inits = inits, parameters, "model.bug",
                 n.chains = 1, n.iter = 2000, 
                 bugs.directory="c:/Program Files/WinBUGS14/"
                 , working.directory=getwd()
                 , debug = F
)
# Parameter posterior distributions
mcmc.out4 <- as.mcmc.list(bugs.out)
summary(mcmc.out4, q = c(0.025, 0.975))

# Nice plots
xyplot(mcmc.out4)
densityplot(mcmc.out4)
#autocorr.plot(mcmc.out2) # not super clear (remember how to interpret)

```

```{r model4-newycut}
# Heteroscedastic 4pl model: variance changes as a function of y ----
# plus one new od observation (unknown conc)
# same data as last chunk: model4-newyimp
# Initial values for variance model change
# Chose reasonable values
inits1 <- c(1, 0.1, 3000, -1, 0.002, 1.2
            , list(c(rep(NA, 36), 8000))
            )
names(inits1) <- c("beta1", "beta2", "beta3", "beta4", "sigma.sq", "theta"
                   , "conc.star"
                   )
# If running more than 1 chain, need a set of initial values for each chain
# Maybe as a function such as adding randomness
# inits2 <- 
# inits3 <- 
# As with other BUGS inputs, format as a list
inits <- list(inits1
              #, inits2, inits3
              )

# Parameters to monitor (only need to monitor betas and sigma this time)
parameters <- c(#"x0", 
                "beta1", "beta2", "beta3", "beta4", "sigma", "theta"
                , "conc.star[37]"
                )
model4cut <- function(){
    for (i in 1:J){
        mu[i] <- beta2 + (beta1 - beta2)/(1 + pow(conc[i] / beta3, beta4))
        # Tau changes
        od[i] ~ dnorm(mu[i], tau[i])
        tau[i] <- 1/var[i]
        # same variance model from previous tutorials 
        var[i]  <- sigma.sq * pow(mu[i], 2*theta)
        # distribution of conc[i]
        conc[i] <- cut(conc.star[i])
        conc.star[i] ~ dunif(0, 60000)
    }
    # Should not allow betas 1-3 to be <= 0 
    # (we know that even at x=0, y ne 0 unless it was transformed)
    # beta1 and beta2 had dnorm(0, 0.001), but a mean of zero is not
    # realistic. Are dgammas OK for beta1 and beta2? Or log-normal?
    # Also, beta1 must be > beta2. Could this constraint be added?
    beta1 ~ dunif(0, 10)
    beta2 ~ dunif(0, 10)
     # reduced the upper limit x10 because would not converge
    beta3 ~ dunif(0, 10000)
    # We know that with an immunometric assay beta4 must be <= 0.     
    # Should it be specified like this?
    beta4 ~ dunif(-10,0)
    # Precision, tau, cannot be negative
    # Would like to monitor sigma rather than tau or var, but advice on priors
    # for variance are more common. And for consistency with previous constant
    # variance models (model2):
    sigma.sq ~ dgamma(0.001,0.001)
    sigma <- sqrt(sigma.sq)
    # Could use this prior for theta
    # theta ~ dunif(0, 5)
    # Or could give stronger prior belief to constant variance (ie. theta = 0)
    # and not exclude possibility of decreasing variance...
    theta ~ dnorm(0, 0.1) 
    # missing value priors
#     mu.x ~ dunif(0, 10000)
#     p.x <- 1/pow(sd.x, 2)
#     sd.x ~ dunif(0, 10)
}
write.model(model4cut)
bugs.out <- bugs(data, inits = inits, parameters, "model.bug",
                 n.chains = 1, n.iter = 2000, 
                 bugs.directory="c:/Program Files/WinBUGS14/"
                 , working.directory=getwd()
                 , debug = T
)
# Parameter posterior distributions
mcmc.out4 <- as.mcmc.list(bugs.out)
summary(mcmc.out4, q = c(0.025, 0.975))

# Nice plots
xyplot(mcmc.out4)
densityplot(mcmc.out4)
#autocorr.plot(mcmc.out2) # not super clear (remember how to interpret)

```

## Summary

In this tutorial, we have ...

